---
title: "Noise-Referenced Stability Feature Selection"
author: "Jake Winiski"
date: "17JAN2026"
output: html_document
---

# Summary #

**Noise-Referenced Stability Feature Selection** is a practical workflow for identifying reliable predictors in small-to-moderate numerical datasets where sample size is limited and interpretability matters. The method combines bootstrap resampling with Random Forest permutation importance and an explicit noise baseline to distinguish true signal from spurious associations. For each bootstrap fit, feature importance is compared against a null model generated by shuffling the target, producing a stability metric—**freq_sig**—that reflects how often a feature meaningfully outperforms noise. Features are selected using a two-stage rule: first filtering by stability, then ranking by mean importance. The selected set is subsequently evaluated using models of increasing flexibility (Elastic Net, GAM, Random Forest) to characterize how explainable the target is under linear, smooth nonlinear, and interaction-driven assumptions. The goal is not to build an optimal predictive model, but to produce an interpretable, data-driven understanding of which features are consistently useful, which are redundant, and how much of the target variation is realistically explainable given the available data.

This workflow is a statistically cautious, engineer-friendly way to ask:

**Which features consistently act like real signal in small, noisy datasets?**

# Synthetic Dataset Description #

This notebook uses a purpose-built synthetic dataset designed to mimic the structure and challenges of real small-to-moderate experimental tabular data. The dataset contains several classes of features that reflect common patterns encountered in applied R&D problems. **Strong_linear_1** and **strong_linear_2** are primary linear drivers of the target with large, stable effects. **Moderate_linear** represents a secondary linear influence with weaker signal. **Nonlinear_1** introduces a smooth nonlinear relationship, while **interaction_a** provides a true interaction effect between latent variables. A group of correlated features—**confound_proxy_1**, **confound_proxy_2**, and **confound_proxy_3**—serve as redundant measurements of an underlying latent factor, illustrating how multiple predictors can be individually useful yet interchangeable. **Interaction_b** is a correlated but non-causal proxy for the true interaction term, demonstrating how spurious importance can arise from feature correlation. Finally, several **pure_noise** variables are included to represent irrelevant measurements with no connection to the target.

The target variable is constructed as a mixture of these linear, nonlinear, interaction, and confounding components with added random noise, producing a realistic blend of explainable and irreducible variation. This structure allows the workflow to demonstrate key behaviors: the stability selection step should identify truly informative features while filtering out noise; redundant proxies should appear as useful but non-unique predictors; nonlinear and interaction terms should be recognized by more flexible models; and coefficient analysis should reveal how linear models distribute weight among correlated variables. Together, the dataset provides a controlled environment for illustrating how the Noise-Referenced Stability Feature Selection workflow distinguishes robust signal from redundancy and noise, and how different model classes interpret the same underlying information.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# =============================================================================
# Libraries
# =============================================================================

# Data manipulation and visualization
library(tidyverse)   # dplyr, ggplot2, tidyr, readr, etc.
library(ggrepel)     # Improved plot labeling

# Modeling and machine learning
library(caret)       # Unified modeling interface
library(glmnet)      # Elastic Net / Lasso / Ridge
library(ranger)      # Random Forest implementation
library(mgcv)        # Generalized Additive Models

# Evaluation metrics
library(Metrics)     # RMSE, MAE, etc.
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
train.df <- read.csv("demo_feature_selection_dataset.csv")
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# =============================================================================
# Block: Noise-Referenced Stability Feature Selection (Random Forest)
# =============================================================================
# Purpose
# - Identify a compact set of features that are consistently informative for
#   predicting the target, using repeated bootstrap fits of a Random Forest and
#   permutation importance.
#
# Method (high level)
# - For each bootstrap replicate:
#     1) Fit a Random Forest on a bootstrap sample of the data and compute
#        permutation importance for each feature (imp_real).
#     2) Fit a second Random Forest on the same bootstrap sample, but with the
#        target shuffled to destroy signal, and compute permutation importance
#        (imp_null).
#     3) Define a per-bootstrap "noise floor" as the 95th percentile of imp_null
#        (null_q95), then compute "excess importance" = imp_real - null_q95.
#
# Key outputs and how to interpret them
# - imp_summary (table):
#     * mean:     Mean permutation importance across bootstraps (higher = more
#                 influential on average).
#     * sd:       Variability of importance across bootstraps (higher = less
#                 stable).
#     * freq_sig: Fraction of bootstraps where a feature's importance exceeds
#                 the null-derived noise floor (higher = more consistently
#                 informative, relative to noise).
#     * cv:       Coefficient of variation for mean importance (sd / |mean|);
#                 useful as a quick stability diagnostic.
#
# Feature selection rule (recommended set)
# - Step 1 (stability gate): retain features with freq_sig >= FREQ_THRESH.
# - Step 2 (ranking): from those, take the top N_FEATURES by mean importance.
# - If the stability gate yields too few features, fall back to selecting the
#   top features by mean importance.
#
# Practical guidance
# - If most features have freq_sig near 1.0, the data likely contains redundant
#   or highly correlated predictors; selection will behave primarily like "top N
#   by mean importance." In that case, the main challenge is redundancy and
#   interpretability (not the presence of signal).
# - Suggested starting values:
#     * FREQ_THRESH: 0.65 (exploratory) to 0.80 (more conservative)
#     * N_FEATURES:  6–10 for small n (~30–50) to keep interpretation manageable
# =============================================================================

# ---- USER SETTINGS (edit these) ----
FREQ_THRESH <- 0.65   # Minimum freq_sig to pass the stability gate (importance above null)
N_FEATURES  <- 8      # Max number of features to return after ranking by mean importance
# -----------------------------------

set.seed(123)

# 1) Define target and split X/y
target_name <- names(train.df)[ncol(train.df)]
y <- train.df[[target_name]]
X <- train.df[, -ncol(train.df), drop = FALSE]

# 2) Drop rows with missing TARGET (required)
keep <- !is.na(y)
X <- X[keep, , drop = FALSE]
y <- y[keep]

# Optional: if you have any missing feature values, you can either drop or impute.
# For now (step-by-step), just drop incomplete rows:
cc <- complete.cases(X)
X <- X[cc, , drop = FALSE]
y <- y[cc]

cat("Rows after NA cleanup:", nrow(X), "\n")

# 3) Defensive cleanup: remove near-zero variance + extreme correlation
nzv <- caret::nearZeroVar(X)
if (length(nzv) > 0) X <- X[, -nzv, drop = FALSE]

cor_mat <- suppressWarnings(cor(X))
cor_mat[is.na(cor_mat)] <- 0
high_cor <- caret::findCorrelation(cor_mat, cutoff = 0.98, exact = TRUE)
if (length(high_cor) > 0) X <- X[, -high_cor, drop = FALSE]

cat("Features after cleanup:", ncol(X), "\n")

# 4) Stability selection parameters
B <- 200          # number of bootstrap fits (increase to 300+ for more stability)
trees <- 1200     # trees per fit
mtry <- max(1, floor(sqrt(ncol(X))))
min_node_size <- 5

# 5) Function: one bootstrap fit importance + null baseline (shuffled y)
one_boot_importance <- function(seed) {
  set.seed(seed)
  idx <- sample.int(nrow(X), size = nrow(X), replace = TRUE)

  fit_real <- ranger(
    x = X[idx, , drop = FALSE],
    y = y[idx],
    num.trees = trees,
    mtry = mtry,
    min.node.size = min_node_size,
    importance = "permutation",
    seed = seed
  )
  imp_real <- fit_real$variable.importance

  # Null fit: same bootstrap rows, shuffled target
  y_shuf <- sample(y[idx], replace = FALSE)
  fit_null <- ranger(
    x = X[idx, , drop = FALSE],
    y = y_shuf,
    num.trees = trees,
    mtry = mtry,
    min.node.size = min_node_size,
    importance = "permutation",
    seed = seed + 1
  )
  imp_null <- fit_null$variable.importance

  # Ensure full vectors in consistent column order
  real_full <- rep(0, ncol(X)); names(real_full) <- colnames(X)
  null_full <- rep(0, ncol(X)); names(null_full) <- colnames(X)

  real_full[names(imp_real)] <- as.numeric(imp_real)
  null_full[names(imp_null)] <- as.numeric(imp_null)

  # Null threshold per bootstrap (noise floor)
  null_q95 <- unname(quantile(null_full, 0.95, na.rm = TRUE))

  list(real = real_full, excess = real_full - null_q95)
}

# 6) Run bootstrap importance many times
seeds <- sample.int(1e7, B)
imp_list <- map(seeds, one_boot_importance)

imp_real_mat <- bind_rows(lapply(imp_list, function(o) as_tibble_row(o$real)))
imp_excess_mat <- bind_rows(lapply(imp_list, function(o) as_tibble_row(o$excess)))

# 7) Summarize stability for each feature (robust to underscores in names)
imp_summary <- imp_real_mat %>%
  summarise(across(
    everything(),
    list(
      mean = ~mean(.x, na.rm = TRUE),
      sd   = ~sd(.x, na.rm = TRUE)
    ),
    .names = "{.col}__{.fn}"
  )) %>%
  bind_cols(
    imp_excess_mat %>%
      summarise(across(
        everything(),
        list(
          freq_sig = ~mean(.x > 0, na.rm = TRUE)  # beats null in that bootstrap
        ),
        .names = "{.col}__{.fn}"
      ))
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = c("feature", ".value"),
    names_pattern = "^(.*)__(mean|sd|freq_sig)$"
  ) %>%
  arrange(desc(mean)) %>%
  mutate(cv = sd / (abs(mean) + 1e-9))

# Quick peek (top features by mean importance)
print(imp_summary %>% slice_head(n = 10))

# 8) Recommended feature set rules (user-controlled at top)
candidates <- imp_summary %>%
  filter(freq_sig >= FREQ_THRESH) %>%
  arrange(desc(mean))

# Safety fallback: if the stability gate yields too few, fall back to top by mean importance
if (nrow(candidates) < 3) {
  candidates <- imp_summary %>% arrange(desc(mean))
}

selected_features <- candidates %>%
  slice_head(n = min(N_FEATURES, nrow(candidates))) %>%
  pull(feature)

cat("\nSelected features (recommended):\n")
print(selected_features)

# 9) Optional diagnostic plot: stability (freq_sig) vs mean importance
imp_summary %>%
  ggplot(aes(x = freq_sig, y = mean, label = feature)) +
  geom_point(size = 2) +
  ggrepel::geom_text_repel(
    size = 3,
    max.overlaps = Inf,
    box.padding = 0.4,
    point.padding = 0.3
  ) +
  scale_x_continuous(limits = c(0, 1.02)) +
  expand_limits(y = max(imp_summary$mean) * 1.15) +
  labs(
    title = "Stability vs Mean Permutation Importance",
    x = "Frequency importance > null (freq_sig)",
    y = "Mean permutation importance"
  ) +
  theme_minimal() +
  theme(
    plot.margin = margin(10, 60, 10, 10),
    axis.title = element_text(size = 11),
    plot.title = element_text(face = "bold")
  )
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# =============================================================================
# Block: Model Training and Explanatory Quality Assessment
# =============================================================================
# Purpose
# - Evaluate how well the selected feature set explains variation in the target
#   using three model classes of increasing flexibility:
#     1) Elastic Net (linear relationships)
#     2) Generalized Additive Model (smooth nonlinear effects)
#     3) Random Forest (nonlinear effects + interactions)
#
# What this block answers
# - “Given the features selected in the previous step, how explainable is the
#    target variable under different modeling assumptions?”
#
# Interpretation guide
# - R² represents the proportion of target variance that can be explained by
#   the selected features under each model structure.
# - Similar performance between Elastic Net and Random Forest suggests that
#   most explainable signal is approximately linear.
# - Meaningfully better performance of the GAM indicates the presence of smooth
#   nonlinear relationships.
# - Substantially better performance of Random Forest (relative to Elastic Net
#   and GAM) suggests that interactions or more complex nonlinear effects are
#   contributing additional explanatory power.
# - Low R² across all models implies that the available features have limited
#   explanatory connection to the target, or that much of the target variance is
#   driven by noise or unmeasured factors.
#
# Important context
# - This evaluation is diagnostic rather than predictive. The goal is to
#   understand the structure of the signal in the data, not to build a final
#   production model.
# - Cross-validation metrics here should be interpreted as indicators of
#   explanatory coupling, not as formal out-of-sample performance guarantees.
# - Results are most useful for guiding next steps such as feature engineering,
#   target refinement, or additional data collection.
#
# Output
# - A comparison table of RMSE, MAE, and R² for each model type.
# - Residual plots showing model fit patterns across repeated CV folds.
# =============================================================================

set.seed(123)

# ---- User settings ----
CV_FOLDS   <- 5
CV_REPEATS <- 20
K_SMOOTH   <- 5   # GAM smooth basis size (small for small n)

# 1) Build modeling dataset using selected features
target_name <- names(train.df)[ncol(train.df)]

dat_model <- train.df %>%
  select(all_of(c(selected_features, target_name))) %>%
  drop_na() %>%
  as.data.frame()

X <- dat_model[, selected_features, drop = FALSE]
y <- dat_model[[target_name]]

cat("Rows used for modeling:", nrow(dat_model), "\n")
cat("Number of features:", ncol(X), "\n")

# 2) Create repeated CV folds (same folds used for all models)
set.seed(123)
folds <- replicate(
  CV_REPEATS,
  createFolds(y, k = CV_FOLDS, returnTrain = TRUE),
  simplify = FALSE
)

# Utility: compute metrics
metrics <- function(truth, pred) {
  tibble(
    RMSE = Metrics::rmse(truth, pred),
    MAE  = Metrics::mae(truth, pred),
    R2   = cor(truth, pred, use = "complete.obs")^2
  )
}

# =============================================================================
# Model A) Elastic Net
# - Captures linear relationships with regularization to control collinearity.
# =============================================================================
ctrl <- trainControl(
  method = "repeatedcv",
  number = CV_FOLDS,
  repeats = CV_REPEATS,
  savePredictions = "final"
)

set.seed(123)
fit_en <- train(
  x = X, y = y,
  method = "glmnet",
  trControl = ctrl,
  tuneLength = 20
)

pred_en <- fit_en$pred %>%
  group_by(rowIndex) %>%
  summarise(pred = mean(pred), obs = first(obs), .groups = "drop")

met_en <- metrics(pred_en$obs, pred_en$pred) %>% mutate(model = "Elastic Net")

# =============================================================================
# Model B) Random Forest
# - Flexible nonlinear model capable of capturing interactions automatically.
# =============================================================================
set.seed(123)
fit_rf <- train(
  x = X, y = y,
  method = "ranger",
  trControl = ctrl,
  tuneLength = 10,
  importance = "permutation"
)

pred_rf <- fit_rf$pred %>%
  group_by(rowIndex) %>%
  summarise(pred = mean(pred), obs = first(obs), .groups = "drop")

met_rf <- metrics(pred_rf$obs, pred_rf$pred) %>% mutate(model = "Random Forest")

# =============================================================================
# Model C) GAM (mgcv)
# - Models smooth nonlinear effects while remaining interpretable.
# - Implemented with manual repeated CV to allow use of mgcv::gam.
# =============================================================================
gam_formula <- as.formula(
  paste(
    target_name, "~",
    paste(sprintf("s(%s, k=%d)", selected_features, K_SMOOTH), collapse = " + ")
  )
)

all_preds_gam <- list()

set.seed(123)
rep_id <- 1
for (rep_folds in folds) {
  fold_id <- 1
  for (train_idx in rep_folds) {
    test_idx <- setdiff(seq_len(nrow(dat_model)), train_idx)

    train_df <- dat_model[train_idx, , drop = FALSE]
    test_df  <- dat_model[test_idx,  , drop = FALSE]

    gam_fit <- mgcv::gam(
      formula = gam_formula,
      data = train_df,
      method = "REML",
      select = TRUE
    )

    p <- predict(gam_fit, newdata = test_df)

    all_preds_gam[[length(all_preds_gam) + 1]] <- tibble(
      rowIndex = test_idx,
      obs = test_df[[target_name]],
      pred = as.numeric(p)
    )

    fold_id <- fold_id + 1
  }
  rep_id <- rep_id + 1
}

pred_gam <- bind_rows(all_preds_gam) %>%
  group_by(rowIndex) %>%
  summarise(pred = mean(pred), obs = first(obs), .groups = "drop")

met_gam <- metrics(pred_gam$obs, pred_gam$pred) %>% mutate(model = "GAM (mgcv)")

# =============================================================================
# Compare model performance
# - Provides a concise summary of explanatory quality across model types.
# =============================================================================
all_metrics <- bind_rows(met_en, met_gam, met_rf) %>%
  select(model, RMSE, MAE, R2)

print(all_metrics)

# Diagnostic visualization: residual behavior by model
resid_df <- bind_rows(
  pred_en  %>% mutate(model = "Elastic Net",     resid = obs - pred),
  pred_gam %>% mutate(model = "GAM (mgcv)",      resid = obs - pred),
  pred_rf  %>% mutate(model = "Random Forest",   resid = obs - pred)
)

ggplot(resid_df, aes(x = pred, y = resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0) +
  facet_wrap(~model, scales = "free_x") +
  theme_minimal() +
  labs(title = "Residuals vs Predicted (Repeated CV)", x = "Predicted", y = "Residual")
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# =============================================================================
# Block: Elastic Net Coefficient Stability Analysis
# =============================================================================
# Purpose
# - Examine which features are consistently used by the Elastic Net model across
#   cross-validation resamples, and how stable their estimated effects are.
#
# Why this matters
# - For small or moderately sized datasets, coefficient behavior from Elastic
#   Net is often more interpretable than permutation importance.
# - Looking at coefficients across many resamples provides insight into:
#     * which features are repeatedly selected,
#     * how large their effects tend to be,
#     * and whether their direction (sign) is consistent.
#
# What this block produces
# - freq_nonzero: proportion of CV resamples where a feature receives a
#                 non-zero coefficient (higher = more consistently selected).
# - coef_mean:    average coefficient value across resamples.
# - coef_sd:      variability of the coefficient across resamples.
# - mean_abs:     average magnitude of the coefficient (effect size regardless
#                 of sign).
# - sign_pos_frac: fraction of non-zero coefficients that are positive
#                  (indicates directional stability).
#
# How to interpret the output
# - Features with high freq_nonzero and large mean_abs are the most reliably
#   influential predictors under a linear model.
# - A large coef_sd relative to coef_mean suggests instability or sensitivity
#   to small changes in the data.
# - sign_pos_frac near 0 or 1 indicates a stable effect direction; values near
#   0.5 imply the direction is inconsistent across resamples.
# - Features with low freq_nonzero are weak or redundant and are often dropped
#   by the regularization process.
#
# Requirements
# - This block assumes an Elastic Net model named `fit_en` has already been
#   trained using caret::train().
# - The objects `selected_features`, `train.df`, and `target_name` must exist
#   from previous blocks.
# =============================================================================

# ---- OPERATOR SETTINGS ----
# Choose which lambda value to use when extracting coefficients:
#   "caret_best" : use the lambda selected by caret as optimal
#   "caret_1se"  : approximate 1-standard-error rule (more conservative)
#   "custom"     : manually specify a lambda value below
LAMBDA_MODE <- "caret_best"
LAMBDA_CUSTOM <- NULL        # e.g., 0.1 (used only if LAMBDA_MODE == "custom")
# ---------------------------

stopifnot(exists("fit_en"), exists("selected_features"))

# Helper to choose lambda
get_lambda_to_use <- function(fit_en, mode, custom = NULL) {
  if (mode == "custom") {
    stopifnot(!is.null(custom))
    return(custom)
  }
  # caret stores tuning results; pick lambda minimizing RMSE across the grid
  if (mode == "caret_best") {
    return(fit_en$bestTune$lambda)
  }
  if (mode == "caret_1se") {
    # Not always available from caret; approximate by picking largest lambda
    # with RMSE within 1 SD of the minimum (from fit_en$results).
    res <- fit_en$results
    # If alpha is tuned too, keep only the best alpha row group
    if ("alpha" %in% names(res) && "alpha" %in% names(fit_en$bestTune)) {
      res <- res %>% filter(alpha == fit_en$bestTune$alpha)
    }
    min_rmse <- min(res$RMSE, na.rm = TRUE)
    # caret also gives RMSESD in results for resampling SD
    if (!"RMSESD" %in% names(res)) {
      warning("RMSESD not found; falling back to caret_best lambda.")
      return(fit_en$bestTune$lambda)
    }
    thresh <- min_rmse + res$RMSESD[which.min(res$RMSE)]
    # largest lambda within 1-SE threshold
    lam <- res %>%
      filter(RMSE <= thresh) %>%
      arrange(desc(lambda)) %>%
      slice(1) %>%
      pull(lambda)
    return(lam)
  }
  stop("LAMBDA_MODE must be one of: 'caret_best', 'caret_1se', 'custom'")
}

lambda_use <- get_lambda_to_use(fit_en, LAMBDA_MODE, LAMBDA_CUSTOM)
cat("Using lambda for coefficient summaries:", lambda_use, "\n")

# =============================================================================
# Reconstruct per-resample coefficients from caret model
# =============================================================================
# Approach
# - caret does not directly store coefficients for each CV split.
# - To analyze coefficient stability, we:
#     1) Identify the training/test splits used by caret,
#     2) Refit glmnet on each split using the same alpha and lambda,
#     3) Extract the resulting coefficients,
#     4) Aggregate them across resamples.
# =============================================================================

# Build modeling frame consistently
target_name <- names(train.df)[ncol(train.df)]
dat_model <- train.df %>%
  select(all_of(c(selected_features, target_name))) %>%
  drop_na() %>%
  as.data.frame()

X_all <- as.matrix(dat_model[, selected_features, drop = FALSE])
y_all <- dat_model[[target_name]]

# Identify unique resamples used by caret
pred_df <- fit_en$pred

# If alpha is tuned, lock to the best alpha (so we're analyzing the selected model family)
best_alpha <- if ("alpha" %in% names(fit_en$bestTune)) fit_en$bestTune$alpha else NULL

if (!is.null(best_alpha) && "alpha" %in% names(pred_df)) {
  pred_df <- pred_df %>% filter(alpha == best_alpha)
}

# Identify all unique resample identifiers
resamples <- sort(unique(pred_df$Resample))

# Helper: get test indices for a resample from caret prediction table
get_test_idx <- function(resample_id) {
  pred_df %>%
    filter(Resample == resample_id) %>%
    pull(rowIndex) %>%
    unique() %>%
    sort()
}

# Refit Elastic Net for each resample and extract coefficients
coef_list <- vector("list", length(resamples))
names(coef_list) <- resamples

set.seed(123)
for (i in seq_along(resamples)) {
  r <- resamples[i]
  test_idx <- get_test_idx(r)
  train_idx <- setdiff(seq_len(nrow(dat_model)), test_idx)

  xtr <- X_all[train_idx, , drop = FALSE]
  ytr <- y_all[train_idx]

  # Determine alpha to use
  alpha_use <- if (!is.null(best_alpha)) best_alpha else 1

  # Fit glmnet on this resample's training data
  fit_fold <- glmnet::glmnet(
    x = xtr,
    y = ytr,
    alpha = alpha_use,
    standardize = TRUE
  )

  # Extract coefficients at chosen lambda
  cvec <- as.matrix(glmnet::coef.glmnet(fit_fold, s = lambda_use))
  cvec <- cvec[-1, , drop = FALSE]  # drop intercept

  coef_list[[i]] <- tibble(
    Resample = r,
    feature = rownames(cvec),
    coef = as.numeric(cvec[, 1])
  )
}

coef_df <- bind_rows(coef_list)

# =============================================================================
# Summarize coefficient stability across resamples
# =============================================================================
coef_summary <- coef_df %>%
  group_by(feature) %>%
  summarise(
    freq_nonzero = mean(coef != 0),
    coef_mean = mean(coef),
    coef_sd = sd(coef),
    mean_abs = mean(abs(coef)),
    sign_pos_frac = ifelse(sum(coef != 0) == 0, NA_real_, mean(coef[coef != 0] > 0)),
    .groups = "drop"
  ) %>%
  arrange(desc(freq_nonzero), desc(mean_abs))

print(coef_summary)

# Optional visualization: most stable and influential coefficients
top_n <- min(12, nrow(coef_summary))

coef_summary %>%
  slice_head(n = top_n) %>%
  ggplot(aes(x = reorder(feature, mean_abs), y = coef_mean)) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Elastic Net coefficient summary (across CV resamples)",
    x = NULL,
    y = "Mean coefficient (standardized fit)"
  )

# Optional diagnostic plot: stability vs magnitude
coef_summary %>%
  ggplot(aes(x = freq_nonzero, y = mean_abs, label = feature)) +
  geom_point() +
  ggrepel::geom_text_repel(size = 3, max.overlaps = Inf) +
  theme_minimal() +
  labs(
    title = "EN coefficient stability vs magnitude",
    x = "Frequency coefficient ≠ 0 across resamples",
    y = "Mean |coefficient| across resamples"
  )
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# =============================================================================
# Block: Random Forest Permutation Importance (Post-Training)
# =============================================================================
# Purpose
# - Rank and interpret features based on how much model performance degrades when
#   each feature is permuted (i.e., when its information is destroyed).
#
# What this block answers
# - “Which features does the trained Random Forest rely on most for accurate
#    predictions, and how large is the impact of each feature on error?”
#
# Two supported importance modes
# - ranger_internal (fast):
#     * Uses permutation importance stored in the trained ranger model
#       (fit_rf$finalModel$variable.importance).
#     * Suitable for quick inspection and exploratory ranking.
#
# - manual (ΔRMSE, slower but transparent):
#     * Computes a baseline RMSE on the modeling dataset.
#     * For each feature, permutes the column N_PERM times, re-predicts, and
#       recomputes RMSE.
#     * Importance is summarized as mean( RMSE_permuted − RMSE_baseline ).
#     * Provides an interpretable “error penalty” in the same units as the target.
#
# How to interpret the output
# - Larger ΔRMSE (manual) or larger internal permutation importance indicates a
#   feature that the model depends on more strongly.
# - Features near zero imply little marginal contribution given the other
#   predictors (often due to redundancy / correlation).
# - A large delta_rmse_sd suggests the importance estimate is noisy; increasing
#   N_PERM improves stability.
#
# Practical guidance
# - Use ranger_internal for speed, then switch to manual for a more defensible,
#   unit-based importance estimate and clearer communication.
# - Start with N_PERM ~ 50–100 for exploratory work; increase (e.g., 200+) when
#   you want stable ranks or plan to report results.
#
# Outputs
# - A ranked table of feature importances.
# - A bar plot showing relative importance across features.
# =============================================================================

# ---- OPERATOR SETTINGS ----
RF_IMPORTANCE_MODE <- "manual"  # options: "manual", "ranger_internal"
N_PERM <- 200                   # repeats per feature for manual importance
# ---------------------------

stopifnot(exists("fit_rf"), exists("dat_model"), exists("selected_features"), exists("target_name"))

# Ensure modeling frame is plain data.frame and aligned columns exist
dat_model <- as.data.frame(dat_model)

X_rf <- dat_model[, selected_features, drop = FALSE]
y_rf <- dat_model[[target_name]]

rmse_vec <- function(truth, pred) sqrt(mean((truth - pred)^2, na.rm = TRUE))

# =============================================================================
# Option 1) ranger internal permutation importance (fast)
# =============================================================================
if (RF_IMPORTANCE_MODE == "ranger_internal") {

  rf_final <- fit_rf$finalModel
  if (is.null(rf_final$variable.importance)) {
    stop("No variable.importance found. Refit RF with importance='permutation'.")
  }

  vi <- tibble(
    feature = names(rf_final$variable.importance),
    importance = as.numeric(rf_final$variable.importance)
  ) %>%
    arrange(desc(importance))

  print(vi)

  ggplot(vi, aes(x = reorder(feature, importance), y = importance)) +
    geom_col() +
    coord_flip() +
    theme_minimal() +
    labs(
      title = "Random Forest importance (ranger internal permutation)",
      x = NULL,
      y = "Permutation importance"
    )
}

# =============================================================================
# Option 2) Manual permutation importance (ΔRMSE on full data)
# =============================================================================
if (RF_IMPORTANCE_MODE == "manual") {

  # Baseline prediction and RMSE using the trained caret model
  base_pred <- as.numeric(predict(fit_rf, newdata = X_rf))
  base_rmse <- rmse_vec(y_rf, base_pred)
  cat("RF baseline RMSE (full data):", format(base_rmse, digits = 8), "\n")

  set.seed(123)
  vi_list <- lapply(selected_features, function(f) {
    deltas <- numeric(N_PERM)

    for (i in seq_len(N_PERM)) {
      Xp <- X_rf
      Xp[[f]] <- sample(Xp[[f]])   # permute feature f
      pp <- as.numeric(predict(fit_rf, newdata = Xp))
      deltas[i] <- rmse_vec(y_rf, pp) - base_rmse
    }

    tibble(
      feature = f,
      delta_rmse_mean = mean(deltas),
      delta_rmse_sd = sd(deltas)
    )
  })

  vi <- bind_rows(vi_list) %>%
    arrange(desc(delta_rmse_mean))

  # Print with more digits so small effects don't look like 0
  print(vi %>% mutate(delta_rmse_mean = signif(delta_rmse_mean, 6),
                      delta_rmse_sd   = signif(delta_rmse_sd, 6)))

  ggplot(vi, aes(x = reorder(feature, delta_rmse_mean), y = delta_rmse_mean)) +
    geom_col() +
    coord_flip() +
    theme_minimal() +
    labs(
      title = paste0("Random Forest permutation importance (manual, N_PERM=", N_PERM, ")"),
      x = NULL,
      y = "Mean ΔRMSE after permuting feature"
    )
}

if (!RF_IMPORTANCE_MODE %in% c("manual", "ranger_internal")) {
  stop("RF_IMPORTANCE_MODE must be one of: 'manual', 'ranger_internal'")
}
```



